---
title: "BIS 15L Midterm 2"
output:
  html_document: 
    theme: spacelab
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(here)
library(naniar)
``` 

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your code should be organized, clean, and run free from errors. Be sure to **add your name** to the author header above. You may use any resources to answer these questions (including each other), but you may not post questions to Open Stacks or external help sites. There are 10 total questions.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean! Your plots should use consistent aesthetics throughout.  

This exam is due by **12:00p on Tuesday, February 22**.  

## Gapminder
For this assignment, we are going to use data from  [gapminder](https://www.gapminder.org/). Gapminder includes information about economics, population, social issues, and life expectancy from countries all over the world. We will use three data sets, so please load all three as separate objects.    

1. population_total.csv  
2. income_per_person_gdppercapita_ppp_inflation_adjusted.csv  
3. life_expectancy_years.csv  

```{r}
population_total <- readr::read_csv("data/population_total.csv")
```
```{r}
glimpse(population_total)
```

```{r}
income_per <- readr::read_csv("data/income_per_person_gdppercapita_ppp_inflation_adjusted.csv")
```
```{r}
glimpse(income_per)
```

```{r}
life_expectancy <- readr::read_csv("data/life_expectancy_years.csv")
```
```{r}
glimpse(life_expectancy)
```

1. (3 points) Once you have an idea of the structure of the data, please make each data set tidy (hint: think back to pivots) and store them as new objects. You will need both the original (wide) and tidy (long) data!  
```{r}
population_long <- population_total %>% 
   pivot_longer(-country,names_to="year", values_to="population_tot", values_drop_na = TRUE)
```
```{r}
income_long <- income_per %>% 
   pivot_longer(-country,names_to="year", values_to="income_per_person", values_drop_na = TRUE)
```
```{r}
life_long <- life_expectancy %>% 
   pivot_longer(-country,names_to="year", values_to="life_expect", values_drop_na = TRUE)
life_long
```

2. (1 point) How many different countries are represented in the data? Provide the total number and their names. Since each data set includes different numbers of countries, you will need to do this for each one. 
```{r}
population_total %>% 
  summarise(country)
```
For our population total data, all the countries are listed here and as it says there are 195 total countries. u

```{r}
life_expectancy %>% 
  summarise(country)
```
In the life expectancy data, there was a total of 187 countries and they are all listed above as well. 

```{r}
income_per %>% 
  summarise(country)
```
For the income per person data, 193 total countries listed above as well. 

## Life Expectancy  

3. (2 points) Let's limit the data to 100 years (1920-2020). For these years, which country has the highest average life expectancy? How about the lowest average life expectancy?  

```{r}
life_long %>% 
  filter(between(year, 1920, 2020)) %>% 
  group_by(country) %>% 
  summarise(average=mean(life_expect, na.rm=T)) %>% 
  arrange(desc(average))
```
Here we can see from 1920-2020 the country with the highest life expectancy average is Andorra with 79.77. 

```{r}
life_long %>% 
  filter(between(year, 1920, 2020)) %>% 
  group_by(country) %>% 
  summarise(average=mean(life_expect, na.rm=T)) %>% 
  arrange(average)
```
Here we can see the country with the lowest average life expectancy is Central African Republic with an age of 41.77. 

4. (3 points) Although we can see which country has the highest life expectancy for the past 100 years, we don't know which countries have changed the most. What are the top 5 countries that have experienced the biggest improvement in life expectancy between 1920-2020? 
```{r}
life_wide <- life_long %>% 
  filter(between(year, 1920, 2020)) %>% 
  pivot_wider(names_from=year,values_from=life_expect)
life_wide
```
```{r}
life_diff <- life_wide %>% 
  mutate(diff=`2020`-`1920`) %>% 
  select(country, diff) %>% 
  arrange(desc(diff))
```
Here we can see the country with the biggest improvement for life expectancy from 1920 to 2020 was with Kuwait with an increase of about 57 years.

5. (3 points) Make a plot that shows the change over the past 100 years for the country with the biggest improvement in life expectancy. Be sure to add appropriate aesthetics to make the plot clean and clear. Once you have made the plot, do a little internet searching and see if you can discover what histodrical event may have contributed to this remarkable change.  
```{r}
kuwait <- life_long %>%
  filter(country == "Kuwait", 1919<year, year<2021)
kuwait
```

```{r}
kuwait%>%
  ggplot(aes(x=as.factor(year),y=life_expect, group=country, color=country))+geom_point()+geom_line()+labs(title="Kuwait Life Expectancy from 1920 to 2020", x="Year", y="Life Expectancy")
```
Here we can see there is a huge jump in life expectancy after around the years 1940-1950. When researching, I found that in the 1930s, Kuwait started getting into the petroleum industry. This brought lots of wealth and economic growth which allowed their life expectancy to grow much longer. 

## Population Growth
6. (3 points) Which 5 countries have had the highest population growth over the past 100 years (1920-2020)?  

```{r}
population_long %>% 
  group_by(country) %>% 
  filter(1919<year,year<2021) %>% 
  summarise(population_growth=(max(population_tot)-min(population_tot))) %>% 
  arrange(desc(population_growth)) %>% 
  head(n=5)
```
Here we can see that the top contries in the past 100 years for population growth were India, China, Indonesia, USA, and Pakistan. 
7. (4 points) Produce a plot that shows the 5 countries that have had the highest population growth over the past 100 years (1920-2020). Which countries appear to have had exponential growth?  
```{r}
population_long %>%
  filter(year<2021, year>1919, country == "India"|country == "China"|country == "Indonesia"|country == "United States"|country == "Pakistan")%>%
  ggplot(aes(x=year, y=population_tot, color=country)) + geom_point() + facet_wrap(~country)+
  labs(title = "Top 5 Countries for Population Growth", x = "Year", y = "Population")
```
It seems like both China and India have exponential growth for population. 

## Income
The units used for income are gross domestic product per person adjusted for differences in purchasing power in international dollars.

8. (4 points) As in the previous questions, which countries have experienced the biggest growth in per person GDP. Show this as a table and then plot the changes for the top 5 countries. With a bit of research, you should be able to explain the dramatic downturns of the wealthiest economies that occurred during the 1980's.  
notes: 
- prepare change in GDP 
- up to us
```{r}
income_per%>%
  select("country", "1920", "2020") %>%
  mutate(diff_income = `2020` - `1920`) %>%
  arrange(desc(diff_income)) %>% 
  head(n=5)
```
```{r}
income_long%>% 
   filter(country %in% c("Qatar", "Luxembourg", "Singapore", "Brunei", "Ireland"),between(year,1920,2020))%>%
  ggplot(aes(x=year, y=income_per_person,  group=country,color=country))+geom_point()+geom_line()+facet_wrap(~country) + labs(title="Top 5 Countries Income from 1920 until 2020", x="Year", y="GDP")
```
During the 1980s, we see a large drop especially with Qatar and Brunei for GDP per capita in these graphs. After doing some research, it seems like this was largely due to the petroleum market. For Brunei, there was a huge peak around 1970 and then declines dramatically in the 1980s. I found that the decrease was due to low petroleum prices as well as production cuts in Brunei. For Qatar, in the 1980s, they had a large decrease due to the Organization of Petroleum Exporting Countries that put quotas on their oil production and overall lower prices. 

9. (3 points) Create three new objects that restrict each data set (life expectancy, population, income) to the years 1920-2020. Hint: I suggest doing this with the long form of your data. Once this is done, merge all three data sets using the code I provide below. You may need to adjust the code depending on how you have named your objects. I called mine `life_expectancy_100`, `population_100`, and `income_100`. For some of you, learning these `joins` will be important for your project.  



```{r}
life_long_100 <- life_long %>% 
  filter(year>="1920" & year<="2020")
population_long_100 <- population_long %>% 
  filter(year>="1920" & year<="2020")
income_long_100 <- income_long %>% 
  filter(year>="1920" & year<="2020")
```

```{r}
gapminder_join <- inner_join(life_long_100, population_long_100, by= c("country", "year"))
gapminder_join <- inner_join(gapminder_join, income_long_100, by= c("country", "year"))
gapminder_join
```

10. (4 points) Use the joined data to perform an analysis of your choice. The analysis should include a comparison between two or more of the variables `life_expectancy`, `population`, or `income.` 

```{r}
gapminder_join%>%
  filter(country %in% c("Qatar", "Luxembourg", "Singapore", "Brunei", "Ireland"))%>%
  ggplot(aes(x=population_tot, y=life_expect, group=country, color=country)) + geom_point()+facet_wrap(~country)+ labs(title="Life Expectancy vs Population for Top 5 GDP per Capita Countries", x="Population", y="Life Expectancy")
```
Here I chose to compare life expectancy and population for the five countries that I found before had the top GDP per capita. 

